---
title: "WarblerOverlap_Lucy_Exploration - Lucy's Warbler Data Exploration"
author: "Kristen Rosamond"
output: html_document
---

2) Data exploration
Install all necessary packages for data exploration: `data.table` package and `tidyverse` package.
```{r}
library(data.table)
library(tidyverse)
```
Load in Lucy's warbler data file as a data frame and summarize data.
```{r}
Lucys_warbler_data<-fread("/ebd_filtered_subset.csv",data.table=FALSE)
summary(Lucys_warbler_data)
```
Table showing the number of Lucy's warbler individuals by country.
```{r}
Lucys_warbler_country_table<-table(Lucys_warbler_data$common_name, Lucys_warbler_data$country)
Lucys_warbler_country_table
```
Table showing the number of Lucy's warbler individuals by county.
```{r}
Lucys_warbler_county_table<-table(Lucys_warbler_data$common_name, Lucys_warbler_data$county)
Lucys_warbler_county_table
```
Bar graph showing countries with recorded Lucy's warbler sightings.
```{r}
ggplot(data=Lucys_warbler_data,mapping = aes(x = country, y = common_name)) +
  labs(title = "Species observations by country", x = "Country",y = "Species") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5),axis.text.y = element_text(angle=90,hjust=-1.5)) +
  geom_bar(stat="identity")
```
  
Bar graph showing states with recorded Lucy's warbler sightings.
```{r}
ggplot(data=Lucys_warbler_data,mapping = aes(x = state, y = common_name)) +
    labs(title = "Species observations by state", x = "State",y = "Species") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) + #reposition title
  theme(axis.text.x = element_text(angle = 90, hjust = 1),axis.text.y = element_text(angle=90,hjust=-1)) + #make x-axis labels vertical
  geom_bar(stat="identity")
```
  
Bar graph showing counties with recorded Lucy's warbler sightings.
```{r}
ggplot(data=Lucys_warbler_data,mapping = aes(x = county, y = common_name)) +
    labs(title = "Species observations by county", x = "County",y = "Species") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) + #reposition title
  theme(axis.text.x = element_text(angle = 90, hjust = 1),axis.text.y = element_text(angle=90,hjust=-.85)) + #make x-axis and y-axis labels vertical and reposition them
  geom_bar(stat="identity")
```
  
Visualize Lucy's warblers distribution on a global scale.
```{r}
world<-map_data("world") #create a data frame of global map data
world_map<-ggplot() + geom_polygon(data = world, aes(x=long, y = lat, group = group)) +
  coord_fixed(1.3)
  labs <- data.frame(
  long = Lucys_warbler_data$longitude,
  lat = Lucys_warbler_data$latitude,
  stringsAsFactors = FALSE
  )
world_map + 
  geom_point(data = Lucys_warbler_data, aes(x = longitude, y = latitude,color=common_name), size = 1, show.legend=FALSE) + #add points, remove legend
  ggtitle("Global Distribution of Lucy's Warblers") + #add title
  theme_classic() #add classic theme (has no gridlines)
```
  
Table showing the number of Lucy's warbler individuals by US state.
```{r}
Lucys_warbler_USA <- subset(Lucys_warbler_data, country_code == "US") #create new data frame with subset of data containing only warblers in USA
Lucys_warbler_state_table<-table(Lucys_warbler_USA$common_name, Lucys_warbler_USA$state)
Lucys_warbler_state_table
```
Visualize Lucy's warblers distribution in the United States.
```{r}
states<-map_data("state") #create data frame of polygons that show state boundaries
world_map<-ggplot() + geom_polygon(data = states, aes(x=long, y = lat, group = group),color="black",fill="white") +
  coord_fixed(1.3)
  labs <- data.frame(
  long = Lucys_warbler_USA$longitude,
  lat = Lucys_warbler_USA$latitude,
  stringsAsFactors = FALSE
  )
world_map + 
  geom_point(data = Lucys_warbler_USA, aes(x = longitude, y = latitude,color=common_name), color="turquoise",size = 1,show.legend=FALSE) + #add points, change color to turquoise, remove legend
  ggtitle("Distribution of Lucy's Warblers in the United States") + #add title
  theme_classic() #add classic theme (has no gridlines)
```
  
Table showing the number of Lucy's warbler individuals by county in Arizona.
```{r}
Lucys_warbler_AZ <- subset(Lucys_warbler_data, state_code == "US-AZ") #create new data frame with subset of data containing only warblers in Arizona
Lucys_warbler_AZ_table<-table(Lucys_warbler_AZ$common_name, Lucys_warbler_AZ$county)
Lucys_warbler_AZ_table
```

Visualize Lucy's warbler individuals in state of Arizona with county boundaries visible. Zoom in on Arizona and look at counties. First get a state data.frame with a single state for the outline of AZ. Then create a counties data.frame of the counties in AZ.
```{r}
az_df <- subset(states, region == "arizona")
```
Plot the state and use `theme_void` to get rid of grid lines, axes, etc.
```{r}
counties <- map_data("county")
az_county <- subset(counties, region == "arizona")
az_base <- ggplot(data = az_df, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")
az_base + theme_void()
```
  
Add points, county lines, and title.
```{r}
az_base + theme_minimal() + 
  geom_polygon(data = az_county, fill = NA, color = "white") +
  geom_polygon(color = "black", fill = NA) +
  geom_point(data = Lucys_warbler_AZ, aes(x = longitude, y = latitude,color=common_name), color="red",size = 1,show.legend=FALSE,inherit.aes=FALSE) + theme_void() + ggtitle("Lucy's warblers in AZ counties")
```